---
title: "PADP_8120_Lab2: Data Manipulation"
output:
  html_document:
    highlight: pygments
    theme: cerulean
---

#Goals:

After this lab you will be able to:
- make plots using base R
- make better plots using ggplot2 and lattice
- embed plots in an R Markdown document

The broader goal of today's lab is to give you a start at making graphics in R. This is a complex topic, and you are not expected to have mastered data visualization by the end of class (or really by the end of the course either); my goal today is to give you a foundation that you can build on going forward. 

Let's practice plots using some world population data. Assuming that your current working directory is set to `Lab3_Visualization`, you can read in the .csv file from the `data` subdirectory. Take a look at the file...

```{r eval=TRUE,echo=FALSE,warning=FALSE,message=FALSE}
worldpop = read.csv('data/wdata.csv')
```

```{r eval=FALSE}
worldpop = read.csv('data/wdata.csv')
head(worldpop)
```

Most of the file names are fairly self-explanatory. `imr` is the infant mortality rate, `tfr` is total fertility rate, `le` is life expectance and `leM` and `leF` are male and female life expectancy. 

##Plotting with base R

R has decent base plotting functions, some of which you learned in the previous lab. Base R plots are quick, and can be a great way to take a quick snapshot:

###Scatterplots

Plot `imr` against `tfr`:

```{r}
plot(imr~tfr,data=worldpop,type = 'p')
```

`type='p'` tells R to generate points; in this case, if you remove this extra command, the plot will not change since R's default will be to make a scatter plot. Other plot types for two numerical vectors include lines (`'l'`) and both (`'b'`). 

We can make changes to a plot in base R within the plot function. `pch` changes the character type, `main` is the title of the plot, and `col` is color. There are several ways you can specify colors in R, including 6-digit hex codes (see http://www.color-hex.com/ and http://www.0to255.com/):

```{r}
plot(imr~tfr,data=worldpop,type = 'p', pch=19,main='IMR vs. TFR',col='#CD0A0A')
```

as well as character strings for colors that R recognizes (see http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf): 

```{r}
plot(imr~tfr,data=worldpop,type = 'p',pch=19,main='IMR vs. TFR',col='darkolivegreen4')
```

You can also add layers like a `points` or `lines` layer. `abline` adds a straight line that you specify. `abline` allows you to add in a simple regression model directly and will then plot the coefficient slope. Let's also change the line width (using `lwd`) and the line type (`lty`):

```{r}
plot(imr~tfr,data=worldpop,type = 'p',pch=19,main='IMR vs. TFR',col='blue')
abline(reg = lm(imr~tfr,data=worldpop),lwd=2,lty=2)
```

1. Make a scatter plot of `imr` againts total population (`pop2012`). Does there appear to be any relationship?

2. Change the color, character type, and size of the points.

3. Add a title and edit the x and y axis labels (if you need help, remember to use the helpfile `?plot`)

###Histograms
The `hist` function plots basic histograms:
```{r}
hist(worldpop$imr)
```

You can set the binwidth using `breaks`:
```{r}
hist(worldpop$imr,breaks = c(0,40,80,120,160))
```

4. Make a histogram for total fertility rate (`tfr`). Give it a good title and axis names, and change make the color of the bars.


##Plotting with ggplot

Base R is fine, but we can do a lot better! Hadley Wickham has produced a package, ggplot2, that makes it very easy to produce publication-quality graphics. The basic different between base plotting and ggplot2 in R is that in base plot, commands--generally--have to be nested within the plot function (as you saw above). In contrast, ggplot2 uses an additive approach that is much like the chaining feature you just used in dplyr. First, you specify a ggplot object that designates the data set to be used and sets basic plot guidelines such as what the x-axis and y-axis variables will be. Then, you add a "geom" object, in this case "geom_point", which tells R that you want to make a scatter plot. Finally, you can add another term to specify the title: 

```{r ggplot, eval=TRUE}
library(ggplot2)
LEplot = ggplot(data=worldpop,aes(x=leF,y=leM)) 
LEplot + geom_point() + ggtitle('Male LE vs. Female LE')
```

The base for any ggplot object plotting male life expectancy versus female life expectancy will then be `ggplot(data=worldpop,aes(x=leF,y=leM))` (we have saved this object as `LEplot`). We can add other attributes as well. Notice that for ggplot, we add each term (using the `+` operator):

```{r ggplot, eval=TRUE}
LEplot + geom_point() + ggtitle('Male LE vs. Female LE') + 
  scale_y_continuous('Male life expectancy',limits=c(40,90)) + 
  scale_x_continuous('Male life expectancy',limits=c(40,90))
```

It's very easy to make a simple changes. Try loading the `ggthemes` package and use the `theme_economist()` command to make your plot look like a plot from The Economist:
```{r ggplot2, eval=TRUE}
library(ggthemes)
LEplot + geom_point() + ggtitle('Male LE vs. Female LE') + 
  scale_y_continuous('Male life expectancy',limits=c(40,90)) + 
  scale_x_continuous('Male life expectancy',limits=c(40,90)) + theme_economist()
```

This might seem needlessly complicated at first, but as your plotting needs become more and more complex, this flexibility become a huge advantage. In particular, since you generate new plots additively, debugging is easy since you can typically pinpoint the exact source of the problem. You can use ggplot2 to make just about any kind of plot imaginable. It also has a great wiki where you can get help and "borrow" great plotting codes: http://docs.ggplot2.org/current/

12. Make a plot of your choice in ggplot



## Basic plot

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point")
```

## adding aesthetics

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr, color = area))
p + layer(geom = "point")
```

## adding aesthetics

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr, color = area, size = pop2012))
p + layer(geom = "point")
```

## adding aesthetics

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = tfr, y = le, color = area, size = pop2012))
p + layer(geom = "point")
```

## Basic plot

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point")
```

## changing the geom

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "line")
```

## changing the geom

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "blank")
```

## Basic plot

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point")
```

## adding a stat layer { .smaller }

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point") + layer(stat = "smooth")
```

## adding a stat layer { .smaller }

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point") + layer(stat = "smooth", method = "loess")
```

## adding a stat layer { .smaller }

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point") + layer(stat = "smooth", method = "lm")
```

## adding a stat layer { .smaller }

```{r}
p <- ggplot(data = world.pop.data, 
            aes(x = le, y = tfr))
p + layer(geom = "point") + layer(stat = "smooth", method = "lm", se = FALSE)
```

## Basic plot { .smaller }

```{r}
p <- ggplot(data = world.pop.data, aes(x = le, y = tfr))
```

## adding faceting { .smaller }

```{r}
p <- ggplot(data = world.pop.data, aes(x = le, y = tfr))
p + layer(geom = "point") + facet_grid(area ~ .)  
```

## adding faceting { .smaller }

```{r}
p <- ggplot(data = world.pop.data, aes(x = le, y = tfr))
p + layer(geom = "point") + facet_grid(. ~ area)  
```

## adding faceting { .smaller }

```{r}
p <- ggplot(data = world.pop.data, aes(x = le, y = tfr))
p + layer(geom = "point") + facet_grid(. ~ area) + layer(stat = "smooth", method = "loess")
```




* * *

## On Your Own

-   Make a bar plot (hint: `geom_bar()`) of total homeruns by team.

-   Let's consider a new variable: batting average, the number of hits (H) divided by the total number of at bats (AB). Create this new variable by 
    subtracting dividing the two columns in the data frame and assigning them to a new 
    object called `BA`.

-   Using the `geom_density()` command with ggplot, make a density plot of BA observations. 

-   Describe the distribution of `BA` in terms of its center, shape, and 
    spread, including any plots you use. What does this tell us about how people 
    feel about their current weight?
    
-   Using numerical summaries and a side-by-side box plot (`geom_boxplot()`), determine if players in the 2000s (2000 to 2009) tend to hit more home runs in a season than players in the 1990s (1990 to 1999). 

-   Find the mean and standard deviation of strikeouts
    `SO` and determine what proportion of the weights are within one 
    standard deviation of the mean.

-   Plot the relationship between strikeouts (`SO`) and homeruns (`HR`). Do players who hit more homeruns appear to strikeout more? 

*This lab is adapted from Andrew Bray's openintro statistics lab (https://github.com/andrewpbray)
