---
title: "PADP_8120_Lab2: Data Manipulation"
output:
  html_document:
    highlight: pygments
    theme: cerulean
---

#Goals:

After this lab you will be able to:
- manipulate data using base R functions and the `dplyr` package
- use five main git operations: init, add, commit, push, pull 
- start a project with RStudio and keep it under version control with git
- submit your homework with git and github

## 

This lab is adapted from materials by:

- Matthew Salganik
- Andrew Bray


## Getting started

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone 
survey of 350,000 people in the United States. As its name implies, the BRFSS 
is designed to identify risk factors in the adult population and report 
emerging health trends. For example, respondents are asked about their diet and 
weekly physical activity, their HIV/AIDS status, possible tobacco use, and even
their level of healthcare coverage. The BRFSS Web site 
([http://www.cdc.gov/brfss](http://www.cdc.gov/brfss)) contains a complete 
description of the survey, including the research questions that motivate the 
study and many interesting results derived from the data.

We will focus on a random sample of 20,000 people from the BRFSS survey 
conducted in 2000. While there are over 200  variables in this data set, we will
work with a small subset.

We begin by loading the data set of 20,000 observations into the R workspace. 
After launching RStudio, enter the following command.

```{r load-data, eval=FALSE}
source("http://www.openintro.org/stat/data/cdc.R")
```

The data set `cdc` that shows up in your workspace is a *data matrix*, with each
row representing a *case* and each column representing a *variable*.  R calls 
this data format a *data frame*, which is a term that will be used throughout 
the labs.

To view the names of the variables, type the command

```{r names, eval=FALSE}
names(cdc)
```

This returns the names `genhlth`, `exerany`, `hlthplan`, `smoke100`, `height`, 
`weight`, `wtdesire`, `age`, and `gender`. Each one of these variables 
corresponds to a question that was asked in the survey.  For example, for 
`genhlth`, respondents were asked to evaluate their general health, responding
either excellent, very good, good, fair or poor. The `exerany` variable 
indicates whether the respondent exercised in the past month (1) or did not (0).
Likewise, `hlthplan` indicates whether the respondent had some form of health 
coverage (1) or did not (0). The `smoke100` variable indicates whether the 
respondent had smoked at least 100 cigarettes in her lifetime. The other 
variables record the respondent's `height` in inches, `weight` in pounds as well
as their desired weight, `wtdesire`, `age` in years, and `gender`.

1.  How many cases are there in this data set?  How many variables?  For each 
    variable, identify its data type (e.g. categorical, discrete).

We can have a look at the first few entries (rows) of our data with the command

```{r head, eval=FALSE}
head(cdc)
```

and similarly we can look at the last few by typing

```{r tail, eval=FALSE}
tail(cdc)
```

You could also look at *all* of the data frame at once by typing its name into 
the console, but that might be unwise here.  We know `cdc` has 20,000 rows, so 
viewing the entire data set would mean flooding your screen.  It's better to 
take small peeks at the data with `head`, `tail` or the subsetting techniques 
that you'll learn in a moment.

## Summaries and tables

The BRFSS questionnaire is a massive trove of information.  A good first step in
any analysis is to distill all of that information into a few summary statistics
and graphics.  As a simple example, the function `summary` returns a numerical 
summary: minimum, first quartile, median, mean, second quartile, and maximum. 
For `weight` this is

```{r summary-weight, eval=FALSE}
summary(cdc$weight)
```

R also has built-in functions to compute summary statistics one by one.  For 
instance, to calculate the mean, median, and variance of `weight`, type 

```{r weight-mean-var-median, eval=FALSE}
mean(cdc$weight) 
var(cdc$weight)
median(cdc$weight)
```

2. Compute the minimum, maximum, and average age of respondents. 

While it makes sense to describe a quantitative variable like `weight` in terms
of these statistics, what about categorical data?  We would instead consider the
sample frequency or relative frequency distribution.  The function `table` does
this for you by counting the number of times each kind of response was given.
For example, to see the number of people who have smoked 100 cigarettes in their
lifetime, type

```{r table-smoke, eval=FALSE}
table(cdc$smoke100)
```

or instead look at the relative frequency distribution by typing

```{r table-smoke-prop, eval=FALSE}
table(cdc$smoke100)/nrow(cdc)
```

Notice how R automatically divides all entries in the table by the number of observations (using the nrow function to produce the number of rows, 20,000, in the cdc data) in the 
command above. This is similar to something we observed in Lab 1,
when we multiplied or divided a vector with a number, R applied that action across 
entries in the vectors. As we see above, this also works for tables. 

Also, in Lab 1, we explored the use of logical attributes (TRUE/FALSE). You can use this functionality to explore data as well. For instance, the following code produces a TRUE response for every respondent who 6 ft (72 inches) or more in height, and then summarizes the number of respondents in each category:

```{r, eval=FALSE}
summary(cdc$height>=72)
```

3. Summarize the number of respondents who weigh more than 175 pounds (hint: use the code above as a template). 

4.  Create a numerical summary for `height` and `age`, and compute the interquartile range for each. Compute the relative frequency distribution for `gender` and `exerany`. How many males are in the sample? What proportion of the sample reports being in excellent health?

The `table` command can be used to tabulate any number of variables that you 
provide.  For example, to examine which participants have smoked across each 
gender, we could use the following.

```{r table-smoke-gender, eval=FALSE}
table(cdc$gender,cdc$smoke100)
```

## How R thinks about data

We mentioned that R stores data in data frames, which you might think of as a 
type of spreadsheet. Each row is a different observation (a different respondent)
and each column is a different variable (the first is `genhlth`, the second 
`exerany` and so on). We can see the size of the data frame next to the object 
name in the workspace or we can type

```{r dim, eval=FALSE}
dim(cdc)
```

which will return the number of rows and columns. Now, if we want to access a 
subset of the full data frame, we can use row-and-column notation. For example,
to see the sixth variable of the 567<sup>th</sup> respondent, use the format

```{r cdc-row567-column6, eval=FALSE}
cdc[567,6]
```

which means we want the element of our data set that is in the 567<sup>th</sup> 
row (meaning the 567<sup>th</sup> person or observation) and the 6<sup>th</sup> 
column (in this case, weight). We know that `weight` is the 6<sup>th</sup> variable
because it is the 6<sup>th</sup> entry in the list of variable names:

```{r names-again, eval=FALSE}
names(cdc)
```

To see the weights for the first 10 respondents we can type

```{r first-10-rows-sixth-column, eval=FALSE}
cdc[1:10,6]
```

In this expression, we have asked just for rows in the range 1 through 10.  R 
uses the `:` to create a range of values, so 1:10 expands to 1, 2, 3, 4, 5, 6, 
7, 8, 9, 10. You can see this by entering

```{r numbers-1to10, eval=FALSE}
1:10
```

Finally, if we want all of the data for the first 10 respondents, type

```{r first-10-rows, eval=FALSE}
cdc[1:10,]
```

By leaving out an index or a range (we didn't type anything between the comma 
and the square bracket), we get all the columns. When starting out in R, this is
a bit counterintuitive. As a rule, we omit the column number to see all columns 
in a data frame. Similarly, if we leave out an index or range for the rows, we 
would access all the observations, not just the 567<sup>th</sup>, or rows 1 
through 10. Try the following to see the weights for all 20,000 respondents fly 
by on your screen

```{r 6th-column, eval=FALSE}
cdc[,6]
```

Recall that column 6 represents respondents' weight, so the command above 
reported all of the weights in the data set. An alternative method to access the
weight data is by referring to the name.  Previously, we typed `names(cdc)` to 
see all the variables contained in the cdc data set. We can use any of the 
variable names to select items in our data set. 

```{r weight, eval=FALSE}
cdc$weight
```

The dollar-sign tells R to look in data frame `cdc` for the column called 
`weight`.  Since that's a single vector, we can subset it with just a single 
index inside square brackets.  We see the weight for the 567<sup>th</sup> 
respondent by typing

```{r weight-567, eval=FALSE}
cdc$weight[567]
```

Similarly, for just the first 10 respondents

```{r weight-first10, eval=FALSE}
cdc$weight[1:10]
```

The command above returns the same result as the `cdc[1:10,6]` command. Both 
row-and-column notation and dollar-sign notation are widely used, which one you 
choose to use depends on your personal preference.

## Subsetting

It's often useful to extract all individuals (cases) in a data set that have 
specific characteristics. We accomplish this through *conditioning* commands. 

First, consider expressions like

```{r true-male, eval=FALSE}
cdc$gender == "m"
```

or

```{r true-over30, eval=FALSE}
cdc$age > 30
```

These commands produce a series of `TRUE` and `FALSE`  values. There is one 
value for each respondent, where `TRUE` indicates that the person was male (via 
the first command) or older than 30 (second command).






## Special values in R

- `NA`: not available, missing

- `NULL`: does not exist, is undefined

- `TRUE`, `T`: logical true. **Logical** is also an object class.

- `FALSE`, `F`: logical false

## Finding special values

| Function  | Meaning              |
| --------- | -------------------- |
| `is.na`   | Is the value `NA`    |
| `is.null` | Is the value `NULL`  |
| `isTRUE`  | Is the value `TRUE`  |
| `!isTRUE` | Is the value `FALSE` |

<br>

```{r}
absent <- NA
is.na(absent)
```


## Logical operators

<div class="columns-2">
`<` less than

`<=` less than or equal

`>` greater than

`>=` greater than or equal

`==` exactly equal to

`!=` not equal to

`!x` not equal to x

`x | y` x or y

`x & y` x and y
</div>

### Subsetting with logicals

```{r}
x <- c(12, 15, 8, 11, 24)
i <- c(F, F, T, F, F)
x[i]
which(x < 10)
x[x < 10] <- 10
x
```

Usefel facts:
```{r}
i <- c(F, F, T, F, F)
sum(i)
mean(i)
x <- c(12, 15, 8, 11, 24)
mean(x > 11)
```

### More on logical subsetting

Suppose we want to extract just the data for the men in the sample, or just for 
those over 30. We can use the R function `subset` to do that for us. For example,
the command

```{r males, eval=FALSE}
mdata <- subset(cdc, cdc$gender == "m")
```

will create a new data set called `mdata` that contains only the men from the 
`cdc` data set. In addition to finding it in your workspace alongside its 
dimensions, you can take a peek at the first several rows as usual

```{r head-males, eval=FALSE}
head(mdata)
```

This new data set contains all the same variables but just under half the rows.
It is also possible to tell R to keep only specific variables, which is a topic
we'll discuss in a future lab. For now, the important thing is that we can carve
up the data based on values of one or more variables.

You can use several of these conditions together with `&` and `|`.
The `&` is read "and" so that

```{r males-and-over30, eval=FALSE}
m_and_over30 <- subset(cdc, gender == "m" & age > 30)
```

will give you the data for men over the age of 30. The `|` character is read 
"or" so that

```{r males-or-over30, eval=FALSE}
m_or_over30 <- subset(cdc, gender == "m" | age > 30)
```

will take people who are men or over the age of 30 (why that's an interesting 
group is hard to say, but right now the mechanics of this are the important 
thing). In principle, you may use as many "and" and "or" clauses as you like 
when forming a subset.

3.  Create a new object called `under23_and_smoke` that contains all observations 
    of respondents under the age of 23 that have smoked 100 cigarettes in their 
    lifetime. Write the command you used to create the new object as the answer
    to this exercise.

## dplyr: A better way to work with data

Hadley Wickham (assistant professor of statistics at Rice, major R guru, and all-around good guy) has a knack for developing packages that make life a lot easier for R users. Two of his best packages are *plyr* and *dplyr*. Dplyr in particular is an excellent way to work with data.

Load (and if necessary install) plyr and dplyr. Note: you should always load plyr before dplyr, as the two packages have some overlapping functions that respond best when plyr is loaded first. 

```{r eval=TRUE,message=FALSE}
#install.packages(c('plyr','dplyr'))
library(dplyr)
```

In Hadley's own words, the dplyr package makes data manipulation fast and easy by:

I. Elucidating the most common data manipulation operations, so that your options are helpfully constrained when thinking about how to tackle a problem.

II. Providing simple functions that correspond to the most common data manipulation verbs, so that you can easily translate your thoughts into code.

III. Using efficient data storage backends, so that you spend as little time waiting for the computer as possible.

At the outset, let's focus on single-table data. These are data that are in a single dataframe or that you might find in a csv file or Excel sheet. Install and load the "Lahman" package, which contains baseball statistical data from Sean Lahman, and load the "batting" data set included in the package. 

```{r load-lahman, eval=TRUE,message=FALSE}
#install.packages('Lahman',repos = "http://cran.us.r-project.org")
library(Lahman)
data(Batting)
```

4. Take a look at the dataset using the head and names functions, as well as other functions that you have learned so far. How many columns and rows are in the data?

At a basic level, dplyr provides you with five tools that you can use to work with a single data table. These tools are:

###filter()

The filter command subsets a data table based upon each observed value. The following code filters out only Seattle Mariner's players:
```{r filter, eval=FALSE}
filter(Batting, teamID=='SEA')
```

5. How would you save this new filtered table as a new object rather than having it simply print off in the console?

You can also filter based upon multiple attributes or for multiple values:

```{r filter2, eval=FALSE}
#only Seattle Mariners players in 2010
filter(Batting, teamID=='SEA' & yearID==2010)
```

```{r filter3, eval=FALSE}
#either Mariners players or San Diego Padres players
filter(Batting, teamID=='SEA'|teamID=='SDN')
```

The filter() function will retain all rows for which the logical query you specify is TRUE. Thus, you can also us != to filter based upon values that do not equal the specified value. 

```{r filter4, eval=FALSE}
#keep all players EXCEPT Seattle Mariners players
filter(Batting, teamID!='SEA')
```

6. Make a new data table that includes ONLY Atlanta Braves players ('ATL' is the abbreviated name in the data table) in year 2000 or later.

###arrange()

The arrange command provides an easy way to sort observations. By default, the smallest value will go at the top; you can invert this using the "-". 

```{r arrange, eval=FALSE}
#sort by number of home runs, lowest to highest
arrange(Batting, HR)
```

```{r arrange2, eval=FALSE}
#sort by number of home runs, highest to lowest
arrange(Batting, -HR)
```

You can also sorty by multiple attributes at the same time, including character values:

```{r arrange3, eval=FALSE}
#sort by number of home runs, highest to lowest, and then by team.
arrange(Batting, -HR, teamID)
```

7. Who are the top three hitters for total number of doubles ('2B') hit in a single season?

###mutate()

The mutate command is used to generate new variables in the data table or to edit existing variables. For instance, we can create a new category (e.g., stolen bases "SB" + home runs ("HR") )

```{r mutate, eval=FALSE}
mutate(Batting, SBHR = SB + HR)
```

or modify an existing variable:

```{r mutate1, eval=FALSE}
mutate(Batting, RBI = RBI + 1000)
```

One important thing to remember is that these changes will not be stored as part of the original object. Thus, you have to assign the mutated data table to an object name (either the same name or a new name):

```{r mutate2, eval=FALSE}
Batting = mutate(Batting, SBHR = SB + HR)
```

8. Generate a new variable of your choosing; be sure that the new variable is saved within the Batting data table (i.e., assign a new object)

###select()

```{r select, eval=FALSE}
#select player id, team id, at bats, and homeruns
select(Batting, playerID,teamID,AB,HR)
```

sometimes, you want to drop one or two columns and keep the rest. It can be incredibly cumbersome to insert all the names that you want to keep. Instead, you can invert the select() function, again using the "-" sign, to drop a specified variable:

```{r select2, eval=FALSE}
#drop homeruns from datatable
select(Batting, -HR)
```

9. Generate a data table that includes every variable except "G" and "G_batting".

###summarise()

Finally, the summarise command can be used to generate summary statistics. For instance, you can compute the mean or median of a given variable. Note that the summarise command is slightly different than some other operations, in that the variable you want to summarise must be called within the function you want. 

```{r summarise, eval=FALSE}
summarise(Batting,mean(HR,na.rm=T))
```

Note also that you need to select na.rm=T within the mean function, otherwise the function will choke on the presence of NA values. Setting na.rm=T tells R to ignore NA values when computing the variable mean. You can request multiple summaries:

```{r summarise2, eval=FALSE}
summarise(Batting,mean(HR,na.rm=T),mean(SB,na.rm=T))
```

```{r summarise3, eval=FALSE}
summarise(Batting,mean(HR,na.rm=T),sd(HR,na.rm=T))
```

or, you can use the summarise_each() function to do the same thing. Notice that the "." stands in for each variable, and then you select na.rm=T within each function. 

```{r summarise4, eval=FALSE}
summarise_each(Batting,funs(min(.,na.rm=T),max(.,na.rm=T)),HR,SB)
```

###group_by()

You might be wondering what the point is of the summarise and summarise_each functions since as-of-yet we have used them to do operations that we can already do quite easily with base functions such as min(), max(), and mean() (e.g., mean(Batting$HR,na.rm=T)). 

The group_by function is a key addition that greatly multiplies the power of dplyr, as it allows us to compute grouped summary values, for instance the maximum number of homeruns hit in a season for each team:

```{r summarise5, eval=TRUE}
summarise(group_by(Batting,teamID),mean(HR,na.rm=T))
```

10. Using the summarise and group_by functions, generate a summary table showing the maximum number of stolen bases (SB) and triples (3B) for each team by a batter in a single season. 

###chaining functions

Finally, perhaps the coolest feature of dplyr is that you can daisy chain functions together. Typically, when working with data you will perform a series of operations; dplyr allows you to link these operations together without needing to generate a series of intermediate objects.

Basically, you do this by starting with the data table (Batting in this case) and then using the "%>%" operator to link functions. Within a chain, you do not need to place the data table name within each operation function:

```{r chaining, eval=FALSE}
Batting %>% filter(AB>400,yearID>1990) %>% group_by(teamID) %>% summarise(max(SB)) 
```

11. Using the chaining technique, identify the minimum number of hits (`H`) made in a single season by a player on each team who had at least 400 at-bats (`AB`) in a single season since year 2000.


#Version Control

## start a project and keep it under version control

Live demo

## start a project and keep it under version control

Questions?

## submit your homework via git

Live demo (http://www.princeton.edu/~mjs3/soc504_s2015/submitting_homework.shtml)

## submit your homework via git

Questions?

## 

Goal check

## 

motivation for next class

